\chapter{Discussion and Conclusion}
\label{cha:discussion-conclusion}

On the circle $S^1$, functions can be decomposed into sine and cosine waves,
i.e., the Fourier basis. This makes the circle a useful setting for studying
spectral bias. Low frequencies describe slowly varying structure, while high
frequencies describe more oscillatory structure. The question is whether the
training dynamics treat these components differently, and whether this
distinction remains meaningful after moving away from the ideal infinite-width
and infinite-data setting.

In the continuum infinite-width limit, the NTK is exactly rotation-invariant,
meaning its associated operator is perfectly diagonalized by the Fourier basis.
Consequently, each Fourier component of the residual evolves independently:
\begin{equation}
	\alpha_p(t)
	=
	\exp(-\lambda_p t)\alpha_p(0),
	\label{eq:discussion-continuum-mode-decay}
\end{equation}
where $\alpha_p(t)$ is the residual coefficient of the $p$-th Fourier basis
function, and $\lambda_p$ is the corresponding kernel eigenvalue. For the ReLU
NTK studied here, the low-frequency eigenvalues are larger, dictating that
low-frequency residual components decay strictly faster. This constitutes the
baseline spectral-bias mechanism
\citep{jacot2018ntk,lee2020wide,rahaman2019spectralbiasneuralnetworks,
	basri2019convergence,bietti2019inductivebiasneuraltangent,
	cao2020understandingspectralbiasdeep}.

The continuum calculation also shows the importance of including a trainable
bias in the hidden layer of the network. With the hidden-bias contribution, odd frequencies
$k\ge 3$ have nonzero eigenvalues. Without this contribution, those frequencies
are null directions of the fixed-kernel dynamics. Thus, the architecture itself
affects whether some frequencies are learned at all by the fixed-kernel
operator.

Restricting to \(\mathcal H_K\) lets us ask how much of the idealized Fourier
description holds once the continuum measure and infinite width are replaced by
finite ones. The frozen and sampled cases turn out to be controlled perturbations
of the continuum picture; the more interesting departure is the evolving kernel at
small width, which does not merely approximate the fixed-kernel dynamics but
reshapes them in a way that strengthens the low-frequency bias.

\section{Answers to the Research Questions}
\label{sec:discussion-answers}

The overarching inquiry of this thesis asked to what extent the Fourier
decomposition of the infinite-width NTK on \(S^1\) persists under finite-sample
and finite-width perturbations. We find that the idealized description survives
as a controlled approximation for frozen and sampled networks on low-frequency
blocks, with an error that decreases in the sample size and width. When the
kernel is allowed to evolve, this description becomes incomplete: in our
experiments, training improves the loss mainly by reweighting spectral strength
toward low frequencies rather than by bringing the eigenspaces into closer
alignment with the Fourier modes, and the alignment can even degrade. The
sub-questions are answered as follows.

\paragraph{Finite-sample perturbation (Q1).}
Replacing the uniform continuum measure on \(S^1\) with \(n\) i.i.d.\ samples
destroys exact Fourier diagonalization. On the low-frequency subspace
\(\mathcal{H}_K\), however, the Fourier structure persists approximately: the
sampled operator deviates from the diagonal Fourier prediction by an error of
order \(O(n^{-1/2})\) on this block, up to logarithmic factors. The frequency
components therefore no longer evolve independently, but the corrected
finite-sample eigenvalues (Proposition~\ref{prop:results-expectation-H}) still
govern the dominant early-time decay of the retained components. As the results
chapter showed, deviations from this prediction can become visible at later
times, once some components have already decayed.

\paragraph{Finite-width perturbation (Q2).}
The effect of finite width depends on whether the tangent kernel is held fixed
or allowed to evolve. With the kernel frozen at initialization---so that
training reduces to kernel regression on the initial tangent features---finite
width acts as a controlled random perturbation of the continuum operator. On
\(\mathcal{H}_K\), the frozen finite-width block concentrates around the
infinite-width Fourier block at the \(O(m^{-1/2})\) rate. When the network is
trained fully, the kernel evolves during training. At small widths, this
evolution reweights the operator toward larger eigenvalues on the low-frequency
blocks, so the effective per-frequency learning rates change during training
beyond what the initial spectrum predicts. This advantage diminishes with
width, where the frozen kernel is already close to the continuum baseline.

\section{The Robustness of the Fixed-Kernel Approximation}
\label{sec:discussion-robustness}

The infinite-width NTK is sometimes treated as a purely asymptotic object whose
relevance to networks of finite width and finite data is unclear, and recent
work has examined finite-width corrections and limitations of the kernel
picture \citep{haninNica2019finite,bordelon2023dynamicsfinitewidthkernel,
	vyas2022limitationsntkunderstandinggeneralization}. The frozen-kernel results
in this thesis speak to a narrower question. On a fixed low-frequency block
\(\mathcal{H}_K\), and with the tangent kernel held fixed, both finite sampling
and finite width perturb the continuum Fourier operator in a controlled way:
the deviation concentrates around the continuum block and decreases at the
\(O(n^{-1/2})\) and \(O(m^{-1/2})\) rates. Within this restricted setting, the
low-frequency spectral structure that underlies spectral bias is not an artifact
of the infinite limit but survives as an approximation whose error we can bound.

The computed continuum eigenvalues decrease quickly with frequency. From
Theorem~\ref{thm:results-continuum-spectrum}, \(\lambda_k = O(k^{-2})\), and
numerically the first few are \(\lambda_0 \approx 0.55\), \(\lambda_1 \approx
0.35\), \(\lambda_2 \approx 0.079\), \(\lambda_3 \approx 0.011\), with the
higher frequencies smaller still. This sets the scale against which the
finite-sample and finite-width perturbations must be read: the bounds control an
absolute error on \(\mathcal H_K\), so a given perturbation leaves a mode's
dynamics intact only if it is small compared to that mode's eigenvalue. The large
low-frequency eigenvalues should therefore tolerate more sampling or width noise
than the small high-frequency ones. For the lowest frequencies we can be
reasonably confident that the continuum prediction still describes the dynamics
at modest sample size or width. At high frequency the eigenvalues are small and
close in size, and the alignment is poor enough that we cannot tell whether an
empirical mode is still following its own eigenvalue or picking up contributions
from neighbouring frequencies. We have not tested this magnitude argument as such,
but it is consistent with the frozen alignment experiments, where low frequencies
are well aligned at small width and higher frequencies need larger widths, and
with the broader finding that high-frequency components are learned later and are
more sensitive to model and data distributions
\citep{rahaman2019spectralbiasneuralnetworks,basri2019convergence,
	basri2020frequencybiasneuralnetworks}.

\section{Kernel Evolution and the Limits of Fixed Dynamics}
\label{sec:discussion-kernel-evolution}

Once the kernel evolves, the fixed-kernel description no longer accounts for the
full dynamics, and our networks sit between the lazy regime, where the tangent
kernel barely moves, and feature learning, where it moves substantially
\citep{chizat2019lazytraining,Mei2018meanfieldview,
	chizat2018globalconvergencegradientdescent}. The evolving kernel helps where it
moves: it improves on the frozen kernel in exactly the small-width range where
the low-frequency block drifts most
(Figure~\ref{fig:results-evolving-block-drift}).

The improvement does not come from better Fourier alignment, which often degrades
at higher frequencies, but from added spectral strength in the lowest blocks, and
the amplitude variants show this is not just tracking where the target places its
mass. Read alongside the fixed-kernel picture, this looks like a second-order
form of the same bias: gradient descent not only learns low frequencies faster
under a fixed spectrum, it also moves the kernel to strengthen those frequencies
further. We have shown this as a correlation with lower loss, not as its
established cause.

Why the added strength favours \(k=0\) and \(k=1\) is where the architecture and choice of
input parameterisation suggests an explanation. With inputs \(x(\varphi)=(\cos\varphi,\sin\varphi)\),
the pre-activation of hidden neuron \(j\) is
\begin{equation}
	w_j^\top x(\varphi) + b_j
	= w_{j_1}\cos\varphi + w_{j_2}\sin\varphi + b_j
	= \|w_j\|\cos(\varphi-\psi_j) + b_j .
	\label{eq:discussion-preactivation-freq-one}
\end{equation}
Frequencies \(k=0\) and \(k=1\)
are therefore present in the parameterisation itself, through the bias and the
input embedding, while the higher harmonics (including the relatively
well-aligned \(k=2\)) appear only indirectly, once the ReLU acts on this
frequency-one pre-activation. The network may then strengthen and keep aligned
exactly the directions it represents most directly, since these need the least
movement to amplify.

This is testable with the present machinery. A target with no constant or
frequency-one component would separate the two readings: if the strength increase
still concentrates on \(k=0\) and \(k=1\), the parameterisation account is
favoured; if it follows the lowest frequency present in the target, then gradient
descent is instead amplifying wherever the residual is largest.
These observations come from a single architecture on \(S^1\) and do not
separate cause from correlation. What they show is that at small width, training
reshapes the kernel in a low-frequency-favouring direction that the fixed-kernel
theory does not capture, with the parameterisation a likely source. A formal
account is left to future work.

\section{Beyond the Uniform Circle Setting}
\label{sec:discussion-beyond-uniform-circle}

On the uniform circle two things line up: the limiting NTK we derived depends
only on the angle between two points, and the uniform distribution treats every
angle the same. When both hold, the kernel acts as a convolution, and the Fourier
modes are its eigenfunctions, each scaled by its own eigenvalue. Both pieces
matter, the kernel's rotation-invariance and the uniformity of the distribution.

A non-uniform input density removes the second piece. As shown in
Section~\ref{subsec:uniform-nonuniform}, the operator is then no longer a
convolution, and the Fourier modes need not be its eigenfunctions. The components
learned first are the leading eigenfunctions of this density-weighted operator
rather than the lowest Fourier frequencies, and
\citet{basri2020frequencybiasneuralnetworks} show that learning also speeds up
where inputs are more concentrated. Spectral bias persists as a preference for
the leading eigenfunctions of the operator, but these need not be the familiar
low frequencies \citep{basri2020frequencybiasneuralnetworks,
	bowman2022spectralbiasoutsidetraining}.

A more realistic input distribution makes the change starker. Real inputs are
not confined to a circle but spread through the input space, and the standard
model of this is a Gaussian on \(\mathbb{R}^d\). The kernel is then no longer a
convolution, and its eigenfunctions are not Fourier modes but Hermite
polynomials, the Gaussian counterpart of sines and cosines
\citep{mei2022generalization}. These form a family of increasing degree, with
degree playing the role frequency did: low-degree polynomials are simple and
slowly varying, a constant, then a line, then a parabola, and the coarse,
low-degree shape of the target is learned first
\citep{cao2020understandingspectralbiasdeep}. We expect the spectral-bias
mechanism to be unchanged, since eigenvalue size sets the decay rates
regardless of the basis. Our proofs, though, would not transfer directly. They
assume a bounded kernel and use Hoeffding, Bernstein, and McDiarmid
inequalities, all of which rely on bounded or controlled-tail summands,
available to us because the circle is compact. A Gaussian measure has unbounded
support, so neither the boundedness nor these inequalities apply directly, and
the bounds would have to be re-derived with tools suited to unbounded inputs.

We do not expect a smooth activation such as sigmoid or \(\tanh\) to take us
outside the framework used here. The Fourier basis is the eigenbasis because the
operator is a convolution, and the derivation in
Appendix~\ref{app:continuum-kernel} shows this follows from the Gaussian
distribution of the parameters rather than from the activation being ReLU. For any
pointwise activation the NTK still depends only on the angle between two inputs, so
the Fourier modes remain its eigenfunctions and the same concentration
inequalities and eigenspace diagnostics should apply. What changes is the
eigenvalues. ReLU is continuous but not differentiable at the origin, and this non-smoothness
is what limits the eigenvalues to a polynomial decay rate; our derivation gives
\(\lambda_k = O(k^{-2})\) in our setting. \citet{bietti2019inductivebiasneuraltangent}
trace this polynomial decay to the non-smooth part of the ReLU kernel, and
\citet{murray2023characterizingspectrumntkpower} show more generally that smoother activations
have faster-decaying NTK spectra. An activation without this non-differentiability,
such as sigmoid or \(\tanh\), would therefore give faster-decaying eigenvalues and
sharpen the bias toward low frequencies. The
fixed-kernel analysis therefore carries over with only the spectrum recomputed.
We are less sure about the evolving-kernel results. These depend on how the kernel
moves during training, which we did not analyse for other activations and would
not expect to behave the same way, so we leave that case open.

\section{Limitations and Future Directions}
\label{sec:discussion-limitations}

The concentration results are fixed-block results. They control the prescribed
low-frequency space \(\mathcal H_K\) for fixed \(K\), so they do not give a
uniform statement over all Fourier modes, nor are they designed to stay sharp
when \(K\) grows with samples (\(n\)) or width (\(m\)). This matters because higher frequencies have
smaller eigenvalues that sit closer together, where the same perturbation has more
effect. The proofs also use entrywise concentration and union bounds, which
capture the correct decreasing trends but discard matrix structure and variance
information, so the bounds are conservative in absolute magnitude. Sharpening them
with operator-norm bounds \citep[e.g.,][]{tropp2015introduction},
variance-sensitive concentration, or tools for U- and V-statistics
\citep{serfling2009approximation} would bring them closer to the observed errors.

The finite-sample theory explains the leading behaviour of the sampled
fixed-kernel dynamics but does not give a sharp time-uniform description of every
component. Once a component has decayed close to zero, small operator errors can
dominate what is plotted, so the late-time deviations we observe are expected but
not quantified per mode.

The largest gap is the evolving kernel. The finite-width theorem controls the
frozen tangent kernel at initialization but not the kernel as it changes during
training, and our results for that regime are entirely empirical. The fixed-kernel
limit is also tied to NTK scaling: in feature-learning or mean-field regimes the
representation moves substantially during training, and a fixed initial spectrum
cannot describe the dynamics \citep{Mei2018meanfieldview,
	chizat2018globalconvergencegradientdescent}. Controlling how the low-frequency
block drifts from initialization, and what that drift does to the residual
dynamics, is the most direct next step.

The alignment diagnostics measure how close the empirical eigenspaces are to the
Fourier subspaces, but alignment is not the same as loss: a well-aligned block
can still be learned slowly if its eigenvalue is small, so these diagnostics
describe the geometry of the operator rather than the convergence directly.

A full description of practical training would need finite sampling, finite width,
and training time treated together, which this work does not attempt. Our results
cover the first two with the kernel frozen; the training-time effect is only
studied empirically, and it is not simply additional noise, since at small width
the kernel drift changes the dynamics in a structured, low-frequency-favouring
way. Combining all three is the broadest open problem left here.

The analysis also points to a concrete intervention. The fixed-kernel dynamics are
a linear system whose per-mode convergence rate is set by the kernel eigenvalues,
and preconditioning is the natural way to flatten that spectrum: rescaling each
Fourier mode by its inverse eigenvalue clusters the effective rates and removes the
frequency-dependent slowdown, as the theory preconditioner does in
Section~\ref{sec:results-preconditioned-sampled-dynamics}. This is available to us
because the operator is known in closed form and diagonalised by a fixed basis, so
the preconditioner can be applied directly to the residual in function space.

A finite-width trained network does not meet either condition. The operator that
sets the rates is the tangent kernel, which has no closed form at finite width and
drifts during training, and gradient descent acts on the parameters rather than on
the function values, so the correction would have to be applied to the
parameter-space gradient through this same evolving, implicitly known kernel.
Preconditioning the rates in that setting, with an operator that is neither fixed
nor explicitly available, is the real difficulty and is left to future work.

\section{Reflection}
\label{sec:discussion-reflection}

This thesis sits within the broader line of work that tries to replace empirical
intuition about deep learning with analysable mathematical models. The Neural
Tangent Kernel is one of the most useful instruments in that line: it turns a
non-convex training problem into a linear one whose behaviour can be
characterised exactly. The value of the present work is as a controlled case
study of where that reduction holds and where it stops: the fixed kernel is a
useful baseline, but not a full account of training once the network learns
features.

\paragraph{Implications, applicability, and stakeholders.}
Spectral bias has practical consequences wherever neural networks represent
functions with fine detail, through the same NTK spectrum studied here.
\citet{wang2022whenpinnsfail} trace the failure of physics-informed neural
networks to fit high-frequency parts of a PDE solution to a gap in the
convergence rates of the different loss terms, and correct it by rescaling the
dynamics with the kernel eigenvalues. This is the same eigenvalue-based correction
used by the preconditioner in
Section~\ref{sec:results-preconditioned-sampled-dynamics}. In computer vision and
graphics, \citet{tancik2020fourierfeatures} and \citet{sitzmann2020siren} show
that coordinate-based networks (networks that map a raw input coordinate, such as a
pixel location or a 3D point, directly to the signal value) blur fine
detail for the same reason, and that
Fourier-feature mappings and periodic activations fix this by reshaping the
effective kernel spectrum so high frequencies are no longer suppressed. The
mechanism this thesis analyses on the circle is the one these methods modify in
practice.

These methods are built and used by scientific-machine-learning practitioners
working on physical simulation, by people building implicit neural
representations for images, shapes, and audio, and by researchers designing ways
to counteract spectral bias.
Knowing why and when a network learns some components before others makes its
behavior more predictable, which matters when these models inform downstream
decisions. This work does not bear directly on the safety, fairness, or societal
deployment of machine-learning systems, but we hope it adds to the understanding of one
specific mechanism. However, a result on the uniform circle is still far from a
guarantee about a deployed model.

\section{Conclusion}
\label{sec:discussion-final-conclusion}

This thesis studied spectral bias in a setting where the ideal dynamics can be
written down explicitly. On the uniform circle the infinite-width NTK is diagonal
in the Fourier basis, each frequency has its own learning rate, and low
frequencies are learned faster because their eigenvalues are larger.

This picture survives finite sampling and frozen finite width on the fixed
low-frequency block: the ideal operator is perturbed, but the perturbation
decreases with sample size and width at the \(O(n^{-1/2})\) and \(O(m^{-1/2})\)
rates. The evolving finite-width experiments mark where the fixed-kernel account
stops. Once the kernel moves during training, the Fourier basis remains useful for
measurement, but the dynamics are also shaped by how the operator itself changes,
and at small width that change reinforces the low-frequency bias rather than
following it passively.

The Fourier basis is therefore exact in the continuum model, approximately valid on
the low-frequency block under finite sampling and frozen finite width, and a
measurement tool for the evolving kernel. Turning that last regime into theory is
the main problem left open by this work.
