\chapter{Discussion}
\label{cha:discussion}

This chapter interprets the results in relation to the research questions. The
main reference point is the continuum infinite-width NTK on \(S^1\), where the
uniform measure makes the limiting kernel rotation-invariant and hence diagonal
in the Fourier basis. The thesis then asks what remains of this simple
frequency-wise picture after three changes: replacing the continuum measure by
finite samples, replacing the infinite-width kernel by a frozen finite-width
kernel, and finally allowing the finite-width kernel to evolve during training.

The discussion is organized around these three regimes. The first part discusses
finite sampling, the second part discusses the role of the trainable bias, and
the third part discusses finite width, including both the frozen and evolving
kernel experiments.

\section{Overall answer to the main research question}
\label{sec:discussion-main-question}

The main research question was whether the clean Fourier-based learning dynamics
of the continuum infinite-width NTK survive when the ideal assumptions are
relaxed. In the ideal setting, each Fourier mode evolves independently and its
decay rate is determined by the corresponding NTK eigenvalue. The question is
what remains of this statement after finite sampling, finite width at
initialization, and finite-width training with an evolving tangent kernel.

In the continuum infinite-width setting, the Fourier decomposition is exact.
Proposition~\ref{prop:results-continuum-ntk} shows that the limiting NTK is
rotation-invariant on \(S^1\), so the associated operator is a convolution
operator. The Fourier modes therefore diagonalize it, and
Theorem~\ref{thm:results-continuum-spectrum} gives the corresponding
eigenvalues. This is the standard NTK spectral mechanism in an explicit setting:
components aligned with larger kernel eigenvalues decay faster
\citep{jacot2018ntk,lee2020wide,cao2020understandingspectralbiasdeep}. Because
the domain and measure are rotationally symmetric, this eigenvalue ordering is
also a frequency ordering
\citep{bietti2019inductivebiasneuraltangent,basri2019convergence,basri2020frequencybiasneuralnetworks}.

The finite-sample and finite-width results show that the Fourier structure
survives in different ways in different regimes. After finite sampling, exact diagonalization
is lost, but the sampled operator remains close to the diagonal Fourier
prediction on fixed low-frequency blocks. At finite width, the frozen tangent
operator at initialization is a random perturbation of the continuum Fourier
block and concentrates around it as width increases. These results give a
controlled perturbation theory around the ideal NTK model.

The part not settled here is the evolving finite-width case. During training,
the tangent operator can change, so the frozen Fourier block structure need not
remain fixed. The finite-width result should therefore be read as a
frozen-kernel baseline, not as a complete theory of finite-width training
\citep{bordelon2023dynamicsfinitewidthkernel,vyas2022limitationsntkunderstandinggeneralization,golikov2022neuraltangentkernelsurvey}.

\section{Effect of finite sampling in the infinite-width setting}
\label{sec:discussion-q1}

The first research question was about finite sampling, i.e. replacing
the uniform measure on \(S^1\) by \(n\) sampled points. In the continuum model,
the NTK operator is diagonal in the Fourier basis. After sampling, the operator
is represented only through its values on \(n\) sampled input points. The
question is whether the retained low-frequency Fourier modes still behave
approximately like eigenvectors of this sampled operator, and whether the
resulting learning rates still match the Fourier eigenvalue prediction.

\subsection{Approximate Fourier structure on finite samples}
\label{subsec:discussion-q1-structure}

In the continuum setting, the Fourier basis diagonalizes the NTK operator
exactly. After sampling, this is no longer true, the sampled Fourier functions
are not exactly orthonormal on the sampled points, and they are not exact
eigenvectors of the sampled operator. The finite-sample results quantify these
two changes through \(G-I_d\), \(H-\Lambda^{(n)}\), and
\(A\Phi-\Phi\Lambda^{(n)}\) (refer section \ref{sec:finite-sample-methodology}).

Theorem~\ref{thm:results-finite-sample-concentration} and
Figure~\ref{fig:results-finite-sample-diagnostics} show that these errors
decrease with \(n\) on a fixed Fourier block. Thus the randomly sampled problem
lies between two idealized population settings: the uniform population operator,
where harmonic modes diagonalize the kernel, and non-uniform population
operators, where the eigendirections become distribution-dependent
\citep{basri2020frequencybiasneuralnetworks,bietti2019inductivebiasneuraltangent,cao2020understandingspectralbiasdeep}.
The contribution here is an explicit finite-sample perturbation statement for a
prescribed Fourier block.

\subsection{Implications for residual dynamics}
\label{subsec:discussion-q1-dynamics}

The sampled dynamics in
Figures~\ref{fig:results-random-sampling-first200} and
\ref{fig:results-random-sampling-highfreq} show how this approximate structure
affects training. Early in training, the residual mode curves follow the
finite-sample diagonal prediction. This indicates that the main rate ordering is
still governed by the corrected eigenvalues \(\lambda_p^{(n)}\).

The reason the curves eventually deviate is visible from the operator
decomposition
\[
	A\Phi
	=
	\Phi\Lambda^{(n)}
	+
	\bigl(A\Phi-\Phi\Lambda^{(n)}\bigr).
\]
The first term is the diagonal prediction. If this were the whole operator
action, then each retained Fourier mode would evolve independently at its
predicted rate. The second term is the part of the sampled operator action that
is not captured by this diagonal prediction.

To see the effect on the dynamics, suppose for the moment that the sampled
residual lies mainly in the retained Fourier span, so that
\[
	r_t \approx \Phi \alpha_t .
\]
Then applying the sampled operator gives
\[
	A r_t
	\approx
	A\Phi\alpha_t
	=
	\Phi\Lambda^{(n)}\alpha_t
	+
	\bigl(A\Phi-\Phi\Lambda^{(n)}\bigr)\alpha_t .
\]
The first term gives the ideal mode-wise decay. The second term is an additional
forcing term coming from the finite-sample error. It can move mass between the
retained Fourier coordinates and can also push the residual away from the
diagonal prediction.

Early in training, the residual components are still large, so the diagonal
term is the dominant part of the motion. This is why the observed curves follow
the predicted rates well at the beginning. Later, after some modes have already
decayed by several orders of magnitude, the remaining error term can become
visible in the normalized curves. A small absolute error can look large once the
quantity being plotted is already very small. This explains why the diagonal
prediction captures the initial decay rates better than the late-time behavior.

\subsection{Preconditioning as a diagnostic}
\label{subsec:discussion-q1-preconditioning}

The preconditioning experiment gives a second test of the same interpretation.
The theory preconditioner uses only the diagonal prediction
\(\Lambda^{(n)}\), while the empirical preconditioner uses the observed block
\(C_n^{(K)}=G^{-1}H\). If both behave similarly, then most of the observed rate
separation is explained by eigenvalue scaling. If the empirical preconditioner
is substantially better, then finite-sample mixing inside the retained block is
important.

Figures~\ref{fig:results-preconditioned-mode-decay},
\ref{fig:results-theory-precond-spectrum}, and
\ref{fig:results-preconditioner-gap} show that the theory preconditioner already
removes most of the rate separation on the retained block. This is notable
because \(P_{\mathrm{th}}\) only rescales the retained Fourier coefficients by
the predicted eigenvalues. It does not use the observed matrix \(C_n^{(K)}\), so
it does not correct the finite-sample rotation or mixing of the retained
Fourier directions.

The empirical preconditioner is closer to the ideal flattened dynamics because
it uses \(C_n^{(K)}\). It therefore corrects both the eigenvalue scaling and the
mixing caused by the particular sampled points. The fact that
\(P_{\mathrm{th}}\) remains close to \(P_{\mathrm{emp}}\) in this experiment
suggests that most of the rate separation is explained by the predicted
Fourier eigenvalues. The remaining difference is due to the sampled operator not
being exactly diagonal in the retained Fourier basis.

This also gives a useful interpretation of the preconditioning experiment. If
the target is known to be supported mainly on a fixed range of low frequencies,
and if the sampled Fourier block is already close to diagonal, then rescaling by
the predicted eigenvalues can make those retained modes converge on more similar
time scales.

\subsection{Answer to Q1}
\label{subsec:discussion-answer-q1}

Finite sampling does not preserve the exact Fourier decomposition, but it does
preserve it approximately on fixed low-frequency blocks. As \(n\) increases, the
sampled Fourier basis becomes closer to orthonormal, and applying \(A\) to the
retained Fourier modes becomes closer to multiplying each mode by its predicted
eigenvalue. This is why the corrected eigenvalues predict the main early-time
decay rates. The remaining late-time deviations come from the small part of
\(A\Phi\) that is not captured by the diagonal prediction
\(\Phi\Lambda^{(n)}\).

\section{Role of the trainable bias}
\label{sec:discussion-bias}

The trainable hidden bias affects which fourier modes are present in the
limiting NTK. Corollary~\ref{cor:results-bias-spectrum} shows that the trainable hidden bias
changes the nullspace of the continuum operator. Without the bias contribution,
the odd modes \(k\ge 3\) have zero eigenvalue. With the bias contribution, these
modes have nonzero but small eigenvalues. Thus the bias determines whether these
modes can be learned at all in the fixed-kernel continuum dynamics.

This is consistent with earlier harmonic analyses of two-layer ReLU networks on
the sphere and circle, where bias-free two-layer models do not express odd
harmonics beyond the first frequency
\citep{basri2019convergence,basri2020frequencybiasneuralnetworks}. In the
present calculation, the same effect appears directly in the NTK decomposition:
removing the trainable hidden-bias contribution sets the odd eigenvalues
\(k\ge 3\) to zero, while adding it makes them nonzero. Thus
Figure~\ref{fig:results-continuum-odd-modes} distinguishes modes that are
present with small eigenvalues from modes that are absent because the
corresponding eigenvalues vanish.

\section{Effect of finite-width kernel in the continuum setting}
\label{sec:discussion-q2}

The second research question concerns finite width while keeping the continuum
input measure fixed on \(S^1\). There are two cases. In the frozen case, the tangent kernel
is fixed at its random initialization. In the evolving case, the tangent kernel
changes during training. The frozen case can be compared directly with the
continuum Fourier operator, while the evolving case must also account for how
training changes the operator itself.

\subsection{Frozen tangent operator as the current baseline}
\label{subsec:discussion-q2-frozen}

The current finite-width result addresses the frozen tangent operator at
initialization. Theorem~\ref{thm:results-finite-width-frozen-concentration}
shows that
\[
	\mathbb E[B_m^{(K)}]=\Lambda_K
\]
and that \(B_m^{(K)}\) concentrates around \(\Lambda_K\) as \(m\) increases.
Thus, on a fixed low-frequency block, finite width acts as a random perturbation
of the continuum Fourier operator.

Figure~\ref{fig:results-finite-width-concentration} supports this conclusion.
The median and \(95\%\) quantile of
\(\|B_m^{(K)}-\Lambda_K\|_{\max}\) both decrease as \(m\) increases. The
theoretical envelope is conservative, but the qualitative trend matches the
concentration statement. This agrees with the role of the infinite-width NTK as
a useful baseline, while also showing why finite-width fluctuations can matter
quantitatively
\citep{jacot2018ntk,lee2020wide,bordelon2023dynamicsfinitewidthkernel,vyas2022limitationsntkunderstandinggeneralization}.

\subsection{Frequency ordering in the alignment diagnostics}
\label{subsec:discussion-q2-alignment-ordering}

The subspace-alignment plots give an eigenspace-level view of the same
finite-width effect. Figures~\ref{fig:results-finite-width-projector-error} and
\ref{fig:results-finite-width-principal-angles} show that the matched empirical
eigenspaces become more aligned with the corresponding Fourier subspaces as
\(m\) increases. This matches the trend predicted by the concentration result for
\(B_m^{(K)}-\Lambda_K\).

The ordering across frequencies is not perfectly monotone. In particular, the
\(k=2\) Fourier plane is better aligned than some lower-frequency blocks in the
projector-error plot. One possible reason is that \(k=0\) and \(k=1\) are structurally special in this
model. The constant mode is one-dimensional and receives a direct contribution
from the trainable-bias part of the NTK. The first frequency is tied to the
coordinate functions in the embedding
\(x(\varphi)=(\cos\varphi,\sin\varphi)\), and it is the only odd frequency
already present in the no-bias kernel:
\[
	\lambda_1^{(\mathrm{nb})}=\frac14,
	\qquad
	\lambda_k^{(\mathrm{nb})}=0
	\quad \text{for odd } k\ge 3.
\]
Thus \(k=0\) and \(k=1\) are not generic representatives of the
higher-frequency pattern.

The main conclusion from the alignment diagnostics is the width trend. For each
displayed frequency, the matched empirical subspace becomes more aligned with
the corresponding Fourier subspace as \(m\) increases.

\subsection{High-frequency sensitivity}
\label{subsec:discussion-q2-relative}

The finite-width error should not only be judged by its absolute size. It also
matters how large this error is compared with the eigenvalues of the Fourier
modes being studied. In the continuum fixed-kernel model, the residual in mode
\(k\) decays at rate \(\lambda_k\). For low frequencies, \(\lambda_k\) is
relatively large, so a small finite-width perturbation is less likely to change
the observed decay. For high frequencies, \(\lambda_k\) is small, so the same
perturbation can be large compared with the decay rate of that mode. From
Theorem~\ref{thm:results-continuum-spectrum},
\[
	\lambda_k = O(k^{-2})
	\qquad (k\to\infty),
\]
for both even and odd frequencies.

The same issue appears at the eigenspace level. For each \(k\ge 1\), the cosine
and sine modes form a two-dimensional Fourier plane. At finite width, the
matched empirical eigenspace can rotate away from this plane. This rotation is
more visible when neighbouring Fourier eigenvalues are close together, which is
more common at higher frequencies. Thus high-frequency components are not only
slower to learn in the continuum model; they are also more sensitive to finite
sampling and finite-width perturbations. This is consistent with empirical
observations that high-frequency components are learned later and are more
sensitive to the data distribution and to perturbations
\citep{rahaman2019spectralbiasneuralnetworks,basri2019convergence,basri2020frequencybiasneuralnetworks}.

\subsection{Interpreting evolving-kernel drift}
\label{subsec:discussion-q2-evolving-drift}

The frozen-width results explain what happens at initialization. The evolving
experiments ask what changes after the network parameters start moving.
The evolving-kernel experiments add an empirical part to the finite-width
answer. The loss curves show a width-dependent crossover: the evolving tangent
kernel improves the final objective at widths \(128\) and \(256\), gives a
smaller improvement at width \(512\), and loses its advantage at width \(1024\).

The joint diagnostics suggest that this behavior is not explained by improved
Fourier geometry. The subspace cosine similarity plots show that several
higher-frequency subspaces become less aligned with their Fourier reference
subspaces during evolving training. In this sense, kernel evolution does not
make the tangent kernel more Fourier-diagonal.

The spectral-strength plots point to a different effect. During evolving
training, the average kernel strength increases in the lowest Fourier blocks,
especially \(k=0\) and \(k=1\), with a smaller increase for \(k=2\). This
matches the mode-wise residual plots: at small width, the modes that improve
most under evolving dynamics are the low-frequency target modes. Thus the
benefit of kernel evolution in these experiments appears to come from increasing
the strength of the operator on the low-frequency blocks, rather than from
preserving the Fourier eigenspaces more accurately.

The drift diagnostic
\[
	\|C_t^{(K)}-C_0^{(K)}\|_F
\]
supports the same width-dependent picture. It measures how much the projected
low-frequency tangent-kernel block moves away from initialization. The drift is
largest at small width and much smaller at large width. At small width, this
larger change is accompanied by increased low-frequency strength and improved
loss decay. At large width, the frozen kernel is already closer to the
continuum Fourier baseline, and the extra change from kernel evolution is
smaller. This helps explain why the evolving-kernel advantage shrinks with
width.

The amplitude-variant experiments further show that the low-frequency strength
increase is not specific to the original target amplitudes. Even when the
non-constant modes have equal amplitudes, or when the larger amplitudes are
assigned to higher frequencies, the evolving kernel still increases cumulative
low-frequency strength. This does not prove a general causal mechanism, but it
does make the spectral-strength interpretation more robust than an explanation
based only on the original target.

\subsection{Answer to Q2}
\label{subsec:discussion-answer-q2}

For the frozen tangent operator, finite width preserves the continuum Fourier
block up to a random perturbation that decreases with \(m\). The concentration
theorem proves this at the matrix level, and the eigenspace diagnostics show the
same trend geometrically: as width increases, the matched empirical subspaces
become closer to the corresponding Fourier frequency subspaces.

For the evolving tangent kernel, the thesis gives an empirical answer rather
than a theorem. Kernel evolution changes the finite-width dynamics in a
width-dependent way. At small width, it improves the final loss and accelerates
the decay of several low-frequency target modes. The diagnostics suggest that
this improvement is associated with increased low-frequency spectral strength,
not with improved Fourier-subspace alignment. At larger width, the frozen kernel
is already a stronger approximation to the continuum Fourier operator, and the
advantage of the evolving kernel decreases or disappears.

Thus the answer to Q2 is mixed. In the frozen setting, the Fourier description
is recovered as width increases. In the evolving setting, the Fourier subspaces
remain useful diagnostics, but the kernel is no longer simply a fixed
Fourier-diagonal object. It changes during training by both redistributing
strength across frequencies and moving some eigenspaces away from the Fourier
reference. A comparable non-asymptotic theory for this evolving regime remains
open.


\section{Limitations}
\label{sec:discussion-limitations}

\subsection{Scope of the concentration results}
\label{subsec:discussion-limitations-concentration}

The concentration results should be read as fixed-block perturbation results.
They control a prescribed low-frequency space \(\mathcal H_K\) with fixed \(K\).
They do not give a uniform statement over all Fourier modes, and the bounds are
not designed to remain sharp as \(K\) grows with \(n\) or \(m\). This is important
because the high-frequency eigenvalues and eigengaps become small, so controlling
higher and higher frequencies would require increasingly accurate operator
approximations.

The numerical constants are also conservative. In the finite-sample theorem, the
proof controls entries of the relevant matrices and then applies union bounds
over the retained block. The bound for \(H-\Lambda^{(n)}\) uses a
bounded-differences argument, and the bound for
\(A\Phi-\Phi\Lambda^{(n)}\) uses a conditional Bernstein estimate followed by a
union bound over sample-mode pairs. In the finite-width theorem, each projected
entry is controlled using a coarse tail envelope before another union bound is
applied. These steps are sufficient to prove concentration, but they discard
matrix structure and variance information. This explains why the theoretical
envelopes in Figures~\ref{fig:results-finite-sample-diagnostics} and
\ref{fig:results-finite-width-concentration} are much larger than the observed
errors.

In the finite-width concentration plot, the theorem contains an unspecified
universal constant \(c>0\). This constant comes from Bernstein's inequality for
averages of centered sub-exponential random variables
\citep{Vershynin_2018a}. In explicit proofs of such inequalities, constants of
order \(1/4\) often appear after converting a \(\psi_1\)-norm bound into a
moment generating function bound. The plotted choice \(c=0.2\) is a conservative
representative value, not an optimized constant for this problem. Changing this
constant shifts the envelope but does not change the predicted dependence on
\(m\).

A sharper analysis would need to use the structure of the objects being
controlled. For finite sampling, \(H\) is a bounded second-order statistic, so
concentration tools for U- or V-statistics may give tighter estimates. For both
finite sampling and finite width, direct matrix concentration in operator norm
would be preferable to entrywise control followed by norm conversion.

\subsection{Limits of the dynamics claims}
\label{subsec:discussion-limitations-dynamics}

The finite-sample dynamics experiments support the diagonal prediction, but the
theory does not give a complete time-uniform residual bound. The experiments
show that the predicted eigenvalues explain the dominant early-time behavior.
At later times, some residual components have already become very small, so even
a small difference between \(A\Phi\) and \(\Phi\Lambda^{(n)}\) can become visible
in relative terms. The current bounds identify the source of this effect, but
they do not give a sharp description of the full residual trajectory for all
times.

The preconditioning experiment is also a diagnostic rather than a proposed
training method. The theory preconditioner tests whether the observed rate
separation is mainly explained by the diagonal eigenvalue scaling. The empirical
preconditioner tests how much is gained by using the actual sampled
low-frequency block. These experiments support the interpretation of the
sampled dynamics, but they do not establish a general preconditioning algorithm
or a generalization guarantee.

\subsection{Limits of the finite-width alignment diagnostics}
\label{subsec:discussion-limitations-alignment}

The projector-error and principal-angle plots are empirical diagnostics of the
frozen finite-width operator. They compare Fourier frequency subspaces with
matched empirical eigenspaces of a discretized operator. This is useful for
checking whether the eigenspaces become more Fourier-like as width increases,
but it is not a separate theorem.

The matching procedure also introduces a practical limitation. For each
frequency \(k\), the empirical eigenspace is chosen by overlap with
\(\mathcal F_k\). This makes the comparison interpretable, but it also means
that the diagnostic depends on the selected discretization grid and on the
eigenspace-matching rule. The most useful conclusion from these plots is the trend
of errors in frequencies as width increases.

\subsection{Frozen versus evolving finite-width kernels}
\label{subsec:discussion-limitations-evolving}

The finite-width theorem concerns only the tangent operator at initialization.
It does not prove a bound for the evolving tangent kernel during training. The
frozen result therefore controls one source of finite-width error, namely random
initialization of the tangent kernel, but not the time-dependent change of the
kernel along the training trajectory.

The evolving-kernel section provides an empirical comparison between frozen and
evolving dynamics. It tracks the loss, mode-wise residual decay,
Fourier-subspace alignment, spectral strength, and block drift. These
diagnostics show a coherent width-dependent pattern, but they do not amount to a
time-dependent concentration theorem. A full theory would need to control the
evolving block \(C_t^{(K)}\), its drift from \(C_0^{(K)}\), and its action on the
current residual.

There is also no joint theorem combining finite sampling, finite width, and
training time. The finite-sample analysis studies the infinite-width sampled
operator. The finite-width theorem studies the continuum frozen operator. In
actual finite training runs, sampling error, finite-width initialization error,
and kernel evolution occur together.

\subsection{Model assumptions}
\label{subsec:discussion-limitations-model}

The analysis relies on the special structure of \(S^1\) with the uniform
measure. In this setting, the limiting kernel is rotation-invariant, so the
corresponding operator is a convolution operator and the Fourier basis
diagonalizes it exactly. Related harmonic descriptions of neural tangent kernels
on the circle or sphere appear in
\citet{basri2019convergence,basri2020frequencybiasneuralnetworks,bietti2019inductivebiasneuraltangent,cao2020understandingspectralbiasdeep}.
This structure is no longer automatic once the input distribution is changed.
For example, under a non-uniform density on the circle, the integral operator is
weighted by the density and is no longer diagonal in the global Fourier basis;
the relevant eigenfunctions become distribution-dependent, as studied by
\citet{basri2020frequencybiasneuralnetworks}. If the input domain is changed,
the appropriate basis also changes and may not be available in closed form.

The architecture and scaling assumptions are also part of the result. The thesis
uses a two-layer ReLU network in NTK scaling. The ReLU activation is important
because its Gaussian averages have explicit arc-cosine formulas
\citep{cho2009kernel}, which makes the kernel and its Fourier coefficients
computable in closed form. The \(1/\sqrt m\) scaling is important because it
places the model in the NTK or lazy-training regime, where the infinite-width
limit is described by a fixed kernel
\citep{jacot2018ntk,lee2020wide,golikov2022neuraltangentkernelsurvey}.
Changing the activation, adding or removing bias terms, changing the
parameterization, or moving to a feature-learning scaling changes the limiting
kernel and therefore changes the spectrum that drives the dynamics. Such
feature-learning regimes are better described by mean-field or related
infinite-width limits
\citep{Mei2018meanfieldview,chizat2018globalconvergencegradientdescent,sirignano2019meanfieldanalysisneural,Rotskoff_2022}.
Thus the Fourier conclusions should be understood as conclusions about this
explicit NTK model, not as architecture-independent statements about neural
networks in general.
