\chapter{Introduction}\label{cha:intro}

Deep neural networks have achieved remarkable empirical success across a wide
range of application domains, most notably in computer vision and natural
language processing. In vision, deep convolutional and residual architectures
have led to dramatic improvements on large-scale benchmarks, establishing deep
learning as the dominant paradigm for visual recognition tasks
\citep{krizhevsky2012imagenet, he2016deep}. Similarly, in natural language
processing, attention-based Transformer architectures introduced by
\citet{vaswani2023attentionneed} have enabled substantial
advances in sequence modeling and representation learning, and now form the
foundation of modern large-scale language models. Over the past decade, this
empirical progress has been accompanied by a steady increase in model size,
depth, and computational scale, often resulting in highly overparameterized
models that generalize well despite their capacity to fit random labels
\citep{zhang2017understandingdeeplearningrequires, Belkin_2019}. Much of this
progress has been driven by empirical experimentation and architectural
intuition rather than by a complete theoretical understanding of the mechanisms
underlying training and generalization in deep neural networks.

Despite their practical success, the mechanisms governing the training dynamics
and convergence behavior of neural networks remain only partially understood,
even in highly simplified settings \citep{jacot2018ntk, Mei2018meanfieldview,
	chizat2018globalconvergencegradientdescent}. From an optimization perspective,
neural network training involves high-dimensional, non-convex objectives whose
geometry depends intricately on architectural choices, initialization, and
optimization dynamics \citep{bottou2018optimization}. Early analyses of neural
network loss landscapes have highlighted the prevalence of saddle points and
complex critical structures, underscoring the difficulty of directly
characterizing training dynamics in parameter space
\citep{choromanska2015losssurfacesmultilayernetworks}. While more recent
theoretical work has made progress in analyzing overparameterized models under
restrictive assumptions such as two-layer networks or specific scaling regimes
these results do not yet provide a unified explanation of neural network
training in general settings
\citep{du2019gradient,arora2019finegrainedanalysisoptimizationgeneralization}. More broadly,
this gap between empirical performance and theoretical understanding has
motivated the study of neural networks through simplified models and asymptotic
limits, where training dynamics can be analyzed more precisely.

\paragraph{Structure of the thesis.}
This thesis aims to contribute to the theoretical study of neural-network
training by focusing on one such simplified setting in which the learning dynamics can
be analyzed explicitly. We begin by introducing the supervised-learning
framework and empirical risk minimization, which provide the basic language for
describing how models are trained from data. We then consider increasingly
expressive model classes, moving from linear models to deep linear networks and
nonlinear neural networks, to show where training dynamics can be characterized
cleanly and where this understanding begins to break down. This progression
motivates the use of simplified theoretical regimes: settings that are
restrictive enough to permit precise analysis, but still rich enough to capture
phenomena observed in neural-network training. After this foundation, we review
existing theoretical approaches for studying wide neural networks. This
discussion naturally leads to spectral bias, the empirical and theoretical
observation that neural networks often learn smooth, low-frequency components
of a target function before more oscillatory, high-frequency components.
Spectral bias is the main object of study in this thesis. The related work then
summarizes existing empirical and theoretical explanations of this phenomenon,
leading to the research gap and the research questions addressed in the
remainder of the thesis.

\section{Problem setup}

\paragraph{The statistical learning framework}
The fundamental objective in supervised learning is to identify a functional
relationship between an input space $\mathcal{X} \subseteq \mathbb{R}^{d}$ and
a target space $\mathcal{Y}$, for instance \(\mathcal Y=\mathbb R^n\) in
regression or \(\mathcal Y=\{1,\dots,C\}\) in classification where $C$ denotes
the number of classes. We
assume the existence of a joint probability distribution $\rho$ over
$\mathcal{X} \times \mathcal{Y}$. Ideally, one seeks a measurable function $f:
	\mathcal{X} \to \mathcal{Y}$, that minimizes the expected risk (or population
risk):
\begin{equation}
	\mathcal{L}_{\rho}(f) = \mathbb{E}_{\rho}\ell(f(x),y)
	\label{eq:pop-risk}
\end{equation}
where $x \in \mathcal X$, $y \in \mathcal Y$ and $\ell: \mathcal{Y} \times \mathcal{Y} \to [0, \infty]$ is a loss function quantifying
the discrepancy between the prediction and the true target.
\citep{vapnik1999overview, shalev2014understanding}
In practice, two fundamental constraints prevent the direct minimization of the
equation \eqref{eq:pop-risk}. First, searching over the space of all possible
measurable functions is computationally and analytically intractable, so we
usually restrict our search to a specific hypothesis class which is a family a
functions $f_\theta$ parameterized by a vector $\theta \in \mathbb{R}^p$. Second, the
underlying distribution $\rho$ is unknown; we only have access to a finite
training dataset $\mathcal{D} = \{(x_i, y_i)\}_{i=1}^n$ of $n$ samples drawn
i.i.d. from $\rho$, where \(x_i \in \mathcal X\) and \(y_i \in \mathcal Y\).

\paragraph{Empirical risk minimization.}
The second constraint leads us to the principle of
empirical risk minimization (ERM). Given a training dataset $\mathcal{D} = \{(x_i,
	y_i)\}_{i=1}^n$, a measurable function $f_\theta$ parameterized by $\theta$, and a loss function
$\ell(\cdot,\cdot)$, ERM seeks parameters that minimize the empirical risk,
\begin{equation}
	\theta^\star \in
	\arg\min_{\theta} \; \frac{1}{n} \sum_{i=1}^n \ell\!\left(f_\theta(x_i),\,
	y_i\right).
\end{equation}
This formulation isolates the key ingredients that determine the behavior of
learning algorithms: the function class induced by the parameterization $\theta
	\mapsto f_\theta$, the geometry of the loss landscape, and the optimization
procedure used to minimize the empirical risk. In modern neural network
training, the empirical risk is typically minimized using gradient-based
methods \citep{bottou2018optimization}, which induce a dynamical system in
parameter space whose properties depend intricately on both the model
architecture and the chosen loss function.

\paragraph{Linear models and gradient descent.}
To build intuition, we first consider the case of models that are linear in inputs and in parameters.
We define a function $f_w(x): \mathbb{R}^d \to \mathbb{R}$, parameterized by $w \in \mathbb{R}^d$
with the prescription $f_w(x) = x^\top w$. We specialize the loss function to the squared
error, $\ell(f(x), y) =  \frac{1}{2}(f_w(x) - y)^2$.
Let $X\in\mathbb{R}^{d\times n}$ denote the data
matrix whose columns are $x_i \in \mathcal X \subseteq \mathbb{R}^d$, and let $y\in\mathbb{R}^n$ denote the vector
of targets. With slight abuse of notation, the ERM problem simplifies to a least-squares objective:
\begin{equation*}
	\arg\min_w \; \mathcal{L}(w)= \frac{1}{2} \sum_{i=1}^{n}(f_w(x_i) - y_i)^2 = \frac{1}{2}\,\|X^\top w-y\|_2^2.
\end{equation*}
This objective is convex and differentiable, with gradient
\begin{equation}
	\nabla_w \mathcal{L}(w)=X(X^\top w-y)=XX^\top w - Xy.
\end{equation}
At a stationary point $w^\star$, the gradient vanishes, yielding the equations
\begin{equation}
	XX^\top w^\star = Xy,
\end{equation}
which characterize the set of global minimizers of $\mathcal{L}$ owing to convexity.
The structure of the minimizers depends on the rank of the empirical covariance
operator $A := XX^\top$. If $A$ is positive definite (equivalently, if $X$ has
full row rank), then the minimizer is unique and given by $w^\star = A^{-1}Xy$.
In contrast, in the overparameterized regime where $\operatorname{rank}(X)<d$,
the matrix $A$ is singular and the least-squares objective admits infinitely
many global minimizers \citep{boyd2004convex, golub2013matrix}. Among these minimizers, the Moore--Penrose pseudoinverse
selects the one with smallest Euclidean norm. Since $A$ is symmetric, its
pseudoinverse $A^{+}$ acts as the inverse on $\mathrm{range}(A)$ and annihilates
$\ker(A)$ \citep{penrose1955generalized, golub2013matrix}, yielding the minimum-norm least-squares solution
\[
	w^\star_{\min} = A^{+}Xy.
\]
Here minimum-norm means that \(w^\star_{\min}\) minimizes \(\|w\|_2\) among all
global minimizers of \(\mathcal L\) \citep{gunasekar2018characterizing}.
Gradient descent with step size $\eta>0$ takes the explicit form
\begin{equation}
	w_{t+1}=w_t-\eta \nabla_w \mathcal{L}(w_t)
	=\left(I-\eta XX^\top\right)w_t+\eta Xy.
\end{equation}
Assuming $w_0=0$, we can write a closed form solution
\begin{equation}
	w_t
	=\eta \sum_{s=0}^{t-1}\left(I-\eta XX^\top\right)^s Xy.
\end{equation}
Let $\lambda_{\max}(A)$ denote the largest eigenvalue of $A$.
If the step size satisfies $\eta \in (0,\,2/\lambda_{\max}(A))$, then for any
least-squares minimizer $w^\star$ the error $e_t := w_t - w^\star$ evolves as
\begin{equation}
	e_{t+1} = (I - \eta A) e_t .
\end{equation}
Since $A$ is symmetric, it admits an eigendecomposition
\[
	A = Q\Lambda Q^\top,
\]
where $Q\in\mathbb{R}^{d\times d}$ is orthogonal ($Q^\top Q = I$) and
$\Lambda=\mathrm{diag}(\lambda_1,\dots,\lambda_d)$ contains the (real) eigenvalues
of $A$. Moreover, $A$ is positive semidefinite because
$v^\top A v = \|X^\top v\|_2^2 \ge 0$ for all $v$, hence $\lambda_i \ge 0$ and in
particular $\lambda_i \in [0,\lambda_{\max}(A)]$.

Starting from the error recursion
\[
	e_{t+1} = (I-\eta A)e_t,
\]
we express the error in the eigenbasis of $A$ by defining $\tilde e_t := Q^\top e_t$.
Left-multiplying by $Q^\top$ and using $e_t = Q\tilde e_t$ gives
\[
	\tilde e_{t+1}
	= Q^\top (I-\eta A) Q \,\tilde e_t
	= \bigl(I - \eta Q^\top A Q\bigr)\tilde e_t
	= \bigl(I - \eta \Lambda\bigr)\tilde e_t.
\]
Since $\Lambda$ is diagonal, this yields the component-wise dynamics
\begin{equation}\label{eq:componentwise-dynamics}
	\tilde e_{t+1,i} = (1-\eta \lambda_i)\tilde e_{t,i}, \qquad i=1,\dots,d.
\end{equation}
If $\eta \in (0,\,2/\lambda_{\max}(A))$, then for any $\lambda_i>0$ we have
$0<\eta\lambda_i \le \eta\lambda_{\max}(A) < 2$, hence
$-1 < 1-\eta\lambda_i < 1$, i.e.\ $|1-\eta\lambda_i|<1$.
Under the stated condition on $\eta$, we have $|1 - \eta \lambda_i| < 1$ for all
$\lambda_i > 0$, implying geometric decay of all components of $e_t$ in
$\mathrm{range}(A)$. In particular, if $A$ is positive definite, then
$w_t \to w^\star$ as $t \to \infty$. More generally, if $w_0 \in \mathrm{range}(A)$
(for example, $w_0 = 0$), then $w_t$ converges to the minimum-norm least-squares
solution $w^\star = A^{+}Xy$.

The eigen-decomposition above shows that gradient descent acts as a linear
dynamical system whose behavior is governed by the spectrum of the empirical
covariance operator $XX^\top$. Directions corresponding to larger eigenvalues
converge more rapidly, while components in the null space of $XX^\top$ remain
unchanged. In the overparameterized regime, where multiple \emph{interpolating
	solutions} exist, i.e., solutions satisfying $X^\top w = y$, gradient descent from
standard initializations converges to a particular interpolating solution,
namely the minimum-norm solution \citep{boyd2004convex, golub2013matrix}. This illustrates how the optimization algorithm
induces an implicit bias even in this simplest linear setting.

This simple example already contains the basic spectral mechanism that will
reappear throughout the thesis: gradient descent does not reduce all components
of the error at the same speed; instead, the decay rate of each component is
determined by the eigenvalue of the operator governing the dynamics. In the
linear model above, these eigendirections are determined by the empirical
covariance matrix \(XX^\top\), so they do not yet have to correspond to
frequencies.

\paragraph{Deep linear network.}
Even in the absence of nonlinear activation functions, introducing an
additional layer changes the parameterization of the model, even though the
resulting input-output map remains linear. This already leads to substantially
more intricate training dynamics~\citep{saxe2014exactsolutionsnonlineardynamics}. Consider a two-layer linear network with
scalar output,
\begin{equation}
	f_{v,W}(x) := v^\top W x,
\end{equation}
where $W \in \mathbb{R}^{m \times d}$, $v \in \mathbb{R}^{m}$, and $m$ denotes the
width of the hidden layer. The least-squares empirical risk is
\begin{equation}
	\mathcal{L}(v,W)
	:= \frac{1}{2}\,\bigl\|X^\top W^\top v - y\bigr\|_2^2 .
\end{equation}
Although the resulting function is linear in the input, the loss is no longer
convex in the parameters $(v,W)$ due to their multiplicative coupling.

Let
\begin{equation}
	r := X^\top W^\top v - y \in \mathbb{R}^n
\end{equation}
denote the residual vector. The gradients of $\mathcal{L}$ are given by
\begin{equation}
	\nabla_v \mathcal{L} = W X\, r,
	\qquad
	\nabla_W \mathcal{L} = v (X r)^\top .
\end{equation}
Similar to the linear case, the parameters of this model are updated
iteratively using gradient descent with step size $\eta > 0$
\begin{equation*}
	v_{k+1} = v_k - \eta \nabla_v \mathcal{L}(v_k, W_k), \qquad
	W_{k+1} = W_k - \eta \nabla_W \mathcal{L}(v_k, W_k).
\end{equation*}
The analysis of learning dynamics is often simplified by considering the infinitesimal
learning rate limit ($\eta \to 0$). In this regime, the discrete updates
converge to a system of coupled ordinary differential equations (ODEs) known as
gradient flow.
\begin{equation}
	\dot v_t = -\nabla_v \mathcal{L}(v_t,W_t)
	= W_t X \bigl(y - X^\top W_t^\top v_t\bigr),
	\quad
	\dot W_t = -\nabla_W \mathcal{L}(v_t,W_t)
	= v_t \bigl(y - X^\top W_t^\top v_t\bigr)^\top X^\top .
\end{equation}
This coupled system consists of a vector-valued and a matrix-valued ordinary
differential equation and is nonlinear in the parameters, despite the underlying
predictor being linear in the input. As a result, the training dynamics no longer
admit a simple spectral characterization as in the one-layer case.

Nevertheless, recent work has shown that deep linear networks admit a more
structured description in suitable asymptotic regimes. In particular, \citet{chizat2022infinitewidthlimitdeeplinear} study gradient flow for deep
linear networks in an infinite-width limit, where the dynamics are described at
the level of a measure-valued evolution. While no closed-form solution of the
finite-dimensional ODE system is obtained, this framework establishes global
well-posedness, convergence to global minimizers, and the emergence of implicit
regularization effects induced by depth \citep{chizat2022infinitewidthlimitdeeplinear}.

\paragraph{Nonlinear networks.}
The analysis of training dynamics becomes more involved once nonlinear
activation functions are introduced. Consider again a two-layer network of the form
\[
	f_{v,W}(x) = v^\top \phi(Wx),
\]
where $\phi:\mathbb{R}\to\mathbb{R}$ is a continuously differentiable activation
function applied elementwise. For the least-squares objective, the continuous-time evolution of the parameters is governed by the gradient flow system:
\begin{equation}
	\dot v_t = \phi(W_tX)(y - \phi(W_tX)^\top v_t),
	\qquad
	\dot W_t = \left( \phi(W_tX) \odot \left( v_t (y - \phi(W_tX)^\top)^\top \right) \right)X^\top,
\end{equation}
where $\odot$ denotes the Hadamard (elementwise) product.
The gradient flow equations involve both the activations $\phi(Wx)$
and their derivatives $\phi'(Wx)$. In particular, the evolution of the
parameters mixes standard matrix products with elementwise nonlinear
operations. As a consequence, the training dynamics of nonlinear networks do not admit a
simple closed-form or spectral description in general.

Nevertheless, the linear example suggests the kind of structure one would like
to recover: an operator-level description in which different components of the
error decay at different rates. For nonlinear neural networks, identifying such
an operator and understanding its spectrum is substantially more difficult,
because the relevant dynamics depend on the evolving parameters of the network.
This is one reason why studying frequency-dependent learning, or spectral bias,
is non-trivial in general neural networks. The remainder of the thesis therefore
focuses on simplified regimes in which an operator governing the dynamics can be
made explicit, allowing the connection between its spectrum and the observed
learning behavior to be analyzed more precisely.

\section{Thesis motivation}
At a high level, this thesis is motivated by the question of how these networks'
training dynamics can be studied in a mathematically tractable setting without
discarding the phenomena one ultimately seeks to understand. The objective is
not to provide a complete description of neural network training in full
generality, but rather to identify a setting in which the dynamics can be
analyzed explicitly.

This motivates the study of analytical regimes in which the dynamics become
tractable. Although such regimes necessarily involve simplifications like
taking large-width limits, they can still provide useful insight into mechanisms that
are otherwise difficult to access in the full non-linear parameter dynamics.
Two prominent examples of such approaches are the Neural Tangent Kernel (NTK) and mean-field
formulations, which offer complementary perspectives on the behavior of wide
neural networks during training.

In this thesis, that regime is a single-hidden-layer ReLU network with inputs
on the unit circle (1-sphere \(S^1\)). In the infinite-width limit, the NTK
defines a deterministic integral operator. For the uniform distribution on the
circle, this operator is diagonalized by the Fourier basis (the orthonormal
basis of trigonometric functions: $\sin(k\theta), \cos(k\theta)$). This gives a
clean reference picture for spectral bias: each Fourier mode (specific
harmonic i.e. frequency $k$) has its own learning rate. The rest of the thesis
studies how this picture changes at finite sample size and finite width.
