\chapter{Use of Generative AI}\label{cha:genai}

In accordance with the TU Delft guidelines on the use of generative AI in
end projects, I disclose here the tools used in preparing this thesis, their
purpose, and the scope of their application. All ideas, research directions,
calculations, and theorems in this work are my own: the derivations were
carried out manually and double-checked with professors who are experts in the
relevant fields, and generative AI was used only as a supporting aid that I
reviewed and verified at every stage. Concretely, ChatGPT~5.3 Thinking was used for
brainstorming and for organizing my thoughts into actionable plans, and ChatGPT
Codex~5.3 assisted with coding tasks such as setting up and running experiments
and debugging. Generative AI also assisted in writing the scripts used to
produce the plots in this thesis; these scripts were subsequently read and
adjusted manually to ensure correctness. For the written text, large language
model tools were used to improve grammar, spelling, and clarity, to smooth
transitions between sections, and to make the document more coherent. My
handwritten derivations were converted into \LaTeX{} with the help of tools such
as Mathpix and LLM-based assistants (ChatGPT, Gemini), which were also used to
improve the consistency of notation throughout; the underlying mathematics is
my own. Finally, NotebookLM was used as a retrieval system to quickly locate
specific claims within the literature for citation; I read all cited papers
myself, and this tool served purely for convenient lookup. In all cases I retain
full creative and intellectual responsibility for the content, results, and
academic integrity of this thesis, and I have verified the accuracy, validity,
and originality of all material included.
